On Adding Pattern Matching to Haskell-based Deeply Embedded Domain Specific Languages

### 2020-10-09 paper/paper.tex
Typo and minor formatting change
Put Fig 2 back on its own page by itself
wibble
Paragraphs in conclusion; for GADTs cite 2006 paper by Vytiniotis, Weirich, Peyton Jones

There are three interconnected foundational parts: \ttt{E}, \ttt{ERep} and
\ttt{ERepTy}. \ttt{E} is the deep embedding of the EDSL (a GADT that encodes
expressions in the DSL language). \ttt{ERep} is a type class which represents
all Haskell types which can be represented in the DSL. \ttt{ERepTy} is a type
family associated to the \ttt{ERep} type class, which represents a "canonical form"
of the given type. This canonical form can be immediately constructed in the EDSL.
Canonical form types crucially include \ttt{Either} and \ttt{(,)}, which
allow all combinations of basic sum types and product types to be encoded, in the
manner described at the beginning of Section~\ref{sec:ADTRep}.

@@

  \begin{wrapfigure}{r}{0.6\linewidth}

@@

% \captionsetup{justification=centering}
\caption{Rewrite rules}
\label{fig:Rewrites}
% \end{wrapfigure}
\end{wrapfigure}

@@

The basis of deep embeddings in functional languages is well understood.
\cite{Elliott:03:CompileDSEL-JFP} explains the basic idea, and \cite{Gill:2014:DSLSynth}
gives an acceessable and more recent account of the area.

@@

In this paper:
\begin{itemize}
  \item We address the goal of translating Haskell pattern matching into an
    EDSL.

    % An example usage of this mechanism is given in Section
    % \ref{sec:CBackend} in the form of a transformation of a subset of Haskell
    % into an EDSL that is then used to generate C code, effectively compiling
    % that subset of Haskell into C.

  % \item A representation of pattern matching via a deep embedding of Church
  %   encodings of products and coproducts presented in Section \ref{sec:Church}.
  %   This encoding relies on having a uniform representation of algebraic data
  %   types.

  \item To provide the uniform representation of algebraic data types mentioned
    in the previous point, we describe a mechanism for automatically encoding
    ADTs as compositions of \verb|(,)|, \verb|Either| and \verb|()| using
    \verb|ERep| and \verb|ERepTy| in Section \ref{sec:ADTRep}.

  \item A GHC plugin is outlined in Section \ref{sec:CorePlugin} which provides automation for the translation of
    Haskell code into these EDSL representations.

\end{itemize}

>>>>

There are three interconnected foundational parts: \ttt{E}, \ttt{ERep} and
\ttt{ERepTy}. \ttt{E} is the deep embedding of the EDSL (a GADT~\cite{Vytiniotis:2006:Simple} that encodes
expressions in the DSL language). \ttt{ERep} is a type class which represents
all Haskell types which can be represented in the DSL. \ttt{ERepTy} is a type
family associated to the \ttt{ERep} type class, which represents a "canonical form"
of the given type. This canonical form can be immediately constructed in the EDSL.
Canonical form types crucially include \ttt{Either} and \ttt{(,)}, which
allow all combinations of basic sum types and product types to be encoded, in the
manner described at the beginning of Section~\ref{sec:ADTRep}.

@@

  % \begin{wrapfigure}{r}{0.59\linewidth}
\begin{figure}

@@

% \captionsetup{justification=centering}
\caption{Rewrite rules}
\label{fig:Rewrites}
% \end{wrapfigure}
% \end{wrapfigure}
\end{figure}

@@

The basis of deep embeddings in functional languages is well understood.
\cite{Elliott:03:CompileDSEL-JFP} explains the basic idea, and \cite{Gill:2014:DSLSynth}
gives an accessable and more recent account of the area.

@@

In this paper, a representation of pattern matching was given in the form of a
GADT called \verb|E|. A GHC plugin is then used to transform code written in a
subset of Haskell containing algebraic data types, pattern matching, tail
recursion and lambdas into values of this GADT. A backend can then use these
\verb|E a| values to interpret the Haskell code, compile it or otherwise
manipulate it.

Additionally, a uniform representation of algebraic data types in terms of
\verb|(,)|, \verb|Either| and \verb|()|, called \verb|ERepTy|, was outlined for
use in the previously mentioned pattern matching representation. This
representation of ADTs is determined entirely by the structure of a type,
allowing the pattern matching mechanism to act on the structure of values.

% In this paper:
% \begin{itemize}
%   \item We address the goal of translating Haskell pattern matching into an
%     EDSL.

%     % An example usage of this mechanism is given in Section
%     % \ref{sec:CBackend} in the form of a transformation of a subset of Haskell
%     % into an EDSL that is then used to generate C code, effectively compiling
%     % that subset of Haskell into C.

%   % \item A representation of pattern matching via a deep embedding of Church
%   %   encodings of products and coproducts presented in Section \ref{sec:Church}.
%   %   This encoding relies on having a uniform representation of algebraic data
%   %   types.

%   \item To provide the uniform representation of algebraic data types mentioned
%     in the previous point, we describe a mechanism for automatically encoding
%     ADTs as compositions of \verb|(,)|, \verb|Either| and \verb|()| using
%     \verb|ERep| and \verb|ERepTy| in Section \ref{sec:ADTRep}.

%   \item A GHC plugin is outlined in Section \ref{sec:CorePlugin} which provides automation for the translation of
%     Haskell code into these EDSL representations.

% \end{itemize}
